---
title: "Fig6C_Substrate_Enrichments_Across_Levels"
author: "QD"
date: "2025-12-22"
output: html_document
---


```{r}
library(igraph)
library(ggraph)
library(tidygraph)
library(ggforce)
library(tidyverse)
library(readxl)
```

## Create node lists for the layout of the network graph
```{r}
# Create node data frame
layer1 <- data.frame(name = paste0("L1_", 1:5), layer = 1)
layer2 <- data.frame(name = paste0("L2_", 1:3), layer = 2)
layer3 <- data.frame(name = paste0("L3_", 1:8), layer = 3)

# Combine all layers
nodes <- rbind(layer1, layer2, layer3)

g <- make_empty_graph(n = 0, directed = TRUE) %>%
  add_vertices(nrow(nodes), name = nodes$name, layer = nodes$layer)

layout <- matrix(NA, nrow = nrow(nodes), ncol = 2)
for (layer in 1:3) {
  idx <- which(nodes$layer == layer)
  x_pos <- seq(-1, 1, length.out = length(idx))
  y_pos <- rep(-layer, length(idx))  # layer 1 at top, layer 3 at bottom
  layout[idx, ] <- cbind(x_pos, y_pos)
}

## Add edges
edge_df <- data.frame(from = c("L1_1", "L1_1", "L1_1", "L2_1", "L2_1", "L2_1", "L2_1", "L2_1", "L2_1", "L2_1", "L2_3"),
                      to   = c("L2_1", "L2_2", "L2_3", "L3_1", "L3_2", "L3_3", "L3_4", "L3_5", "L3_6", "L3_7", "L3_8"))

# "L1_5", "L2_4" , "L2_4" , "L3_9"
# Step 2: Flatten and transpose for igraph
edge_vec <- as.vector(t(edge_df[, c("from", "to")]))

# Step 3: Add edges
g <- add_edges(g, edge_vec)


```

## Calculate the enrichments scores (NES) from the GSEA across the different substrate levels
```{r}
library(fgsea)
pathways_FAD_1 <- list()

for (group in levels(glycan_annotations_final_cleaned_long_GSEA$FUNCTION_AT_DESTINATION_1)){
  print(group)
  tmp_df <- glycan_annotations_final_cleaned_long_GSEA %>% dplyr::filter(FUNCTION_AT_DESTINATION_1==group)
  cazymes <- tmp_df %>% select(Subfamily) %>% pull()
  pathways_FAD_1[group] <- group
  pathways_FAD_1[[length(pathways_FAD_1)]] <- cazymes
}

## Now the fold change table. Should also filter out cazymes that have no annotation in the pathways of the respective grouping

results_associations_FAD_1 <- siamcat_NW_results_associations %>% filter(cazyme_fam %in% glycan_annotations_final_cleaned_long_GSEA$Subfamily)
ranks <- results_associations_FAD_1$fc
names(ranks) <- results_associations_FAD_1$cazyme_fam
head(ranks)
barplot(sort(ranks, decreasing = T))

## Perform the actual testing

set.seed(1000)
fgseaRes_W_NW_FAD1 <- fgseaMultilevel(pathways_FAD_1, ranks, minSize=1, maxSize = 500)
head(fgseaRes_W_NW_FAD1[order(padj, -abs(NES)), ], n=10)
plotEnrichment(pathways_FAD_1[["GAG"]], ranks)
plotEnrichment(pathways_FAD_1[["DF"]], ranks)
plotEnrichment(pathways_FAD_1[["PG"]], ranks)
plotEnrichment(pathways_FAD_1[["Mucin"]], ranks)

## Test at different glycan annotation level (FAD2)

test <- glycan_annotations_final_cleaned_long %>% select(FUNCTION_AT_DESTINATION_2,Subfamily) %>% drop_na() %>% filter(!FUNCTION_AT_DESTINATION_2 == "Unknown") %>% group_by(FUNCTION_AT_DESTINATION_2) %>% distinct() %>% group_by(FUNCTION_AT_DESTINATION_2) ## Unknown is removed as this is a non-informative substrate level.
test$FUNCTION_AT_DESTINATION_2 <- as.factor(test$FUNCTION_AT_DESTINATION_2)
test$Subfamily <- gsub("([A-Z])_","\\1",test$Subfamily)

## Now convert into list of pathways with proper format for GSEA testing

pathways_FAD_2 <- list()

for (group in levels(test$FUNCTION_AT_DESTINATION_2)){
  print(group)
  tmp_df <- test %>% dplyr::filter(FUNCTION_AT_DESTINATION_2==group)
  cazymes <- tmp_df %>% select(Subfamily) %>% pull()
  pathways_FAD_2[group] <- group
  pathways_FAD_2[[length(pathways_FAD_2)]] <- cazymes
}

## Now the fold change table. Should also filter out cazymes that have no annotation in the pathways of the respective grouping, as these cannot be included in the GSEA anyway.

results_associations_FAD_2 <- siamcat_NW_results_associations %>% filter(cazyme_fam %in% test$Subfamily)
ranks <- results_associations_FAD_2$fc
names(ranks) <- results_associations_FAD_2$cazyme_fam
head(ranks)
barplot(sort(ranks, decreasing = T))

## Perform the actual testing

set.seed(1000)
fgseaRes_W_NW_FAD2 <- fgseaMultilevel(pathways_FAD_2, ranks, minSize=1, maxSize = 500, nproc = 1) ## For some reason the nproc = 1 is necessary for the results to be fully reproducible
head(fgseaRes_W_NW_FAD2[order(padj, -abs(NES)), ], n=10)

## Test at different glycan annotation level (FAD3)
test <- glycan_annotations_final_cleaned_long %>% select(FUNCTION_AT_DESTINATION_3,Subfamily) %>% drop_na() %>% filter(!FUNCTION_AT_DESTINATION_3 == "Unknown") %>% group_by(FUNCTION_AT_DESTINATION_3) %>% distinct() %>% group_by(FUNCTION_AT_DESTINATION_3) ## Unknown is removed as this is a non-informative substrate level.
test$FUNCTION_AT_DESTINATION_3 <- as.factor(test$FUNCTION_AT_DESTINATION_3)
test$Subfamily <- gsub("([A-Z])_","\\1",test$Subfamily)

## Now convert into list of pathways with proper format for GSEA testing

pathways_FAD_3 <- list()

for (group in levels(test$FUNCTION_AT_DESTINATION_3)){
  print(group)
  tmp_df <- test %>% dplyr::filter(FUNCTION_AT_DESTINATION_3==group)
  cazymes <- tmp_df %>% select(Subfamily) %>% pull()
  pathways_FAD_3[group] <- group
  pathways_FAD_3[[length(pathways_FAD_3)]] <- cazymes
}

## Now the fold change table. Should also filter out cazymes that have no annotation in the pathways of the respective grouping, as these cannot be included in the GSEA anyway.

results_associations_FAD_3 <- siamcat_NW_results_associations %>% filter(cazyme_fam %in% test$Subfamily)
ranks <- results_associations_FAD_3$fc
names(ranks) <- results_associations_FAD_3$cazyme_fam
head(ranks)
barplot(sort(ranks, decreasing = T))

## Perform the actual testing

set.seed(1000)
fgseaRes_W_NW_FAD3 <- fgseaMultilevel(pathways_FAD_3, ranks, minSize=1, maxSize = 500, nproc = 1) ## For some reason the nproc = 1 is necessary for the results to be fully reproducible
head(fgseaRes_W_NW_FAD3[order(padj, -abs(NES)), ], n=10)
plotEnrichment(pathways_FAD_3[["GAG"]], ranks)
```

## Get enrichment scores (NES) from the GSEA so the color gradient can be plotted as such
```{r}
FAD_1_WNW_fgsea_results <- head(fgseaRes_W_NW_FAD1[order(padj, -abs(NES)), ], n=20) %>% filter(pathway != "Other") ## Should consider to remove glycogen here 2, since only 2 CAZymes in there.
FAD_2_WNW_fgsea_results <- head(fgseaRes_W_NW_FAD2[order(padj, -abs(NES)), ], n=20) %>% filter(pathway != "PG") ## Since PG is already in FAD_1
FAD_3_WNW_fgsea_results <- head(fgseaRes_W_NW_FAD3[order(padj, -abs(NES)), ], n=20) %>% filter(pathway != "PG" & size > 10) ## Since PG is already in FAD_1 and only retain nodes with > 10 CAZymes annotated for visualization purposes. Discuss this with Georg though.
complete_fgsea_results_WNW <- rbind(FAD_1_WNW_fgsea_results, FAD_2_WNW_fgsea_results, FAD_3_WNW_fgsea_results)

complete_fgsea_results_WNW <- complete_fgsea_results_WNW %>% mutate(pathway = recode(pathway, "GAG" = "Glycosaminoglycan","DF" = "Dietary Fiber", "PG" = "Peptidoglycan", "Resistant_oligosaccharides" = "Resistant oligosaccharides", "Resistant_starch" = "Resistant starch", "Mannans_and_Heteromannans" = "Mannans and \nheteromannans", "Amylose_and_Amylopectin" = "Amylose and \namylopectin", "NSP" = "Non-starch \npolysaccharides"))
```

## Now add all the glycan names instead of the L1 etc
```{r}
# Get current names
current_names <- V(g)$name

# Create a named vector mapping old names to new names
rename_map <- c("L1_1" = "Dietary Fiber", "L1_2" = "Glycosaminoglycan", "L1_3" = "Mucin", "L1_4" = "Glycogen", "L1_5" = "Peptidoglycan", "L2_1" = "Non-starch \npolysaccharides", "L2_2" = "Resistant oligosaccharides", "L2_3" = "Resistant starch", "L2_4" = "Peptidoglycan", "L3_1" = "Cellulose", "L3_2" = "Pectin", "L3_3" = "Glycoprotein", "L3_4" = "Gum", "L3_5" = "Mannans and \nheteromannans", "L3_6" = "Hemicellulose", "L3_7" = "Chitin", "L3_8" = "Amylose and \namylopectin", "L3_9" = "Peptidoglycan")

# Rename nodes by replacing old names with new ones where applicable
new_names <- ifelse(current_names %in% names(rename_map),
                    rename_map[current_names],
                    current_names)

# Assign back to the graph
V(g)$name <- new_names

substrate_significance <- data.frame(
  name = complete_fgsea_results_WNW$pathway,
  Value = complete_fgsea_results_WNW$padj, ## p-value
  EnrichmentStatus = complete_fgsea_results_WNW$NES) ## This is the NES from the FGSEA

# Step 1: Add the values as a vertex attribute in your graph

# Match the node names and assign corresponding values (or NA if no match)
matched <- match(V(g)$name, substrate_significance$name)
V(g)$value_label <- substrate_significance$Value[matched]
V(g)$enrichment_status <- substrate_significance$EnrichmentStatus[matched]
V(g)$is_significant <- !is.na(V(g)$value_label) & V(g)$value_label < 0.05 ## To have a significance label
```


## Convert to tidygraph object and compute layout
```{r}
tg <- as_tbl_graph(g)

layout_df <- tg %>%
  activate(nodes) %>%
  as_tibble() %>%
  group_by(layer) %>%
  mutate(
    x = seq(-1, 1, length.out = n()),
    y = -1 * (layer - 1) * 3  # Increase spacing between layers (e.g., 3 units apart)
  ) %>%
  ungroup()

layout_df <- as_tibble(tg) %>%
  group_by(layer) %>%
  mutate(
    x = seq(-1, 1, length.out = n()) + 1.5,  # shift right
    y = -1.1 * (layer - 1) ## The lower the -x (so closer to) *, the more the white space is reduced
  ) %>%
  ungroup()

coords <- layout_df[, c("x", "y")]
tg <- tg %>% activate(nodes) %>% mutate(x = coords$x, y = coords$y, enrichment_status = enrichment_status)

```

## Postion the labels
```{r}
label_positions <- layout_df %>%
  group_by(layer) %>%
  summarise(x = mean(x), y = unique(y)) %>%
  mutate(y = y - 0.2)
```

## Get different type of edges
```{r}
layout_df <- layout_df %>% mutate(original_node_name = nodes$name)

make_elbow_edges <- function(edge_df, layout_df, vertical_step = 1) {
  edge_df %>%
    left_join(layout_df, by = c("from" = "original_node_name")) %>%
    dplyr::rename(x_from = x, y_from = y) %>%
    left_join(layout_df, by = c("to" = "original_node_name")) %>%
    dplyr::rename(x_to = x, y_to = y) %>%
    rowwise() %>%
    do({
      #mid_y <- (.$y_from + .$y_to) / 2  # halfway or consistent elbow height
      mid_y <- .$y_to + 0.5 ## To adapt the height of where the horizontal line split occurs
      
      tibble(
        x = c(.$x_from, .$x_from, .$x_to),
        y = c(.$y_from, mid_y, mid_y),
        xend = c(.$x_from, .$x_to, .$x_to),
        yend = c(mid_y, mid_y, .$y_to)
      )
    }) %>%
    bind_rows()
}

## This may have to be debugged for some reason I don't understand yet.

edge_segments <- make_elbow_edges(edge_df, layout_df)
```

## Layout that is more adapted to the ultimate publication
```{r}
network_substrate_graphW_NW <- ggraph(tg, layout = "manual", x = V(g)$x, y = V(g)$y) +
  geom_mark_rect(aes(x = x, y = y, group = factor(layer)),
                 expand = unit(22, "pt"), color = "white", fill = "lightgrey", alpha = 0.2) +
  #geom_edge_link(edge_width = 0.6, alpha = 0.5, color = "black") + ## Have to play a bit with the edge_with still to get this to work properly
  geom_segment(data = edge_segments, aes(x = x, y = y, xend = xend, yend = yend), color = "gray50", linewidth = 0.6) +
  geom_node_point(data = function(x) dplyr::filter(x, is_significant), shape = 21, size = 8.5, stroke = 0.9, fill = NA, color = "grey20") + ## Draw the ring around the significant nodes
  geom_node_point(aes(color = enrichment_status), size = 8) +
  scale_color_gradient2(low = "#3B6FB6", mid = "lightgrey", high = "#D41646", midpoint = 0, na.value = "lightgrey", name = "NES") +
  # Node names on top
  #geom_node_text(aes(label = name), nudge_y = 0.4, size = 2) + 
  # Node names below
  geom_node_text(aes(label = name), nudge_y = -0.28, size = 1.8) + 
  # Values inside nodes, black color
  geom_node_text(aes(label = ifelse(value_label < 0.05, format(value_label, digits = 2, scientific = TRUE), "")), color = "black", size = 1.5) +
  theme_void() +
  theme(legend.title = element_text(size = 6), legend.text  = element_text(size = 5), legend.key.size = unit(0.65, "lines")) +
  theme(legend.position = c(0.5, 1.15), legend.justification = "center", legend.direction = "horizontal", plot.margin = margin(t = 32,  r = 15, b = 50, l = 15)) + # x = center, y = slightly above top
  coord_cartesian(clip = "off")
```

## Now have the same for CRC, so Main Figure 6C
```{r}
## FAD1 testing

results_associations_FAD_1 <- crc_meta_random_effect_study_results_associations %>% filter(cazyme_fam %in% test$Subfamily)
ranks <- results_associations_FAD_1$fc
names(ranks) <- results_associations_FAD_1$cazyme_fam
head(ranks)
barplot(sort(ranks, decreasing = T))

## Perform the actual testing

set.seed(1000)
fgseaRes_CRC <- fgseaMultilevel(pathways_FAD_1, ranks, minSize=1, maxSize = 500, nproc = 1)

##FAD2

results_associations_FAD_2 <- crc_meta_random_effect_study_results_associations %>% filter(cazyme_fam %in% test$Subfamily)
ranks <- results_associations_FAD_2$fc
names(ranks) <- results_associations_FAD_2$cazyme_fam
head(ranks)
barplot(sort(ranks, decreasing = T))

## Perform the actual testing at FAD2

set.seed(1000)
fgseaRes_CRC_FAD_2 <- fgseaMultilevel(pathways_FAD_2, ranks, minSize=1, maxSize = 500, nproc = 1)

## FAD3

results_associations_FAD_3 <- crc_meta_random_effect_study_results_associations %>% filter(cazyme_fam %in% test$Subfamily)
ranks <- results_associations_FAD_3$fc
names(ranks) <- results_associations_FAD_3$cazyme_fam
head(ranks)
barplot(sort(ranks, decreasing = T))

## Perform the actual testing at FAD3

set.seed(1000)
fgseaRes_CRC_FAD_3 <- fgseaMultilevel(pathways_FAD_3, ranks, minSize=1, maxSize = 500, nproc = 1) ## For some reason the nproc = 1 is necessary for the results to be fully reproducible
```

```{r}
## Get enrichment scores (NES) from the GSEA so the color gradient can be plotted as such
FAD_1_CRC_fgsea_results <- head(fgseaRes_CRC[order(padj, -abs(NES)), ], n=20) %>% filter(pathway != "Other") ## Should consider to remove glycogen here 2, since only 2 CAZymes in there.
FAD_2_CRC_fgsea_results <- head(fgseaRes_CRC_FAD_2[order(padj, -abs(NES)), ], n=20) %>% filter(pathway != "PG") ## Since PG is already in FAD_1
FAD_3_CRC_fgsea_results <- head(fgseaRes_CRC_FAD_3[order(padj, -abs(NES)), ], n=20) %>% filter(pathway != "PG" & size > 10) ## Since PG is already in FAD_1 and only retain nodes with > 10 CAZymes annotated for visualization purposes. Discuss this with Georg though.
complete_fgsea_results_crc <- rbind(FAD_1_CRC_fgsea_results, FAD_2_CRC_fgsea_results, FAD_3_CRC_fgsea_results)

complete_fgsea_results_crc <- complete_fgsea_results_crc %>% mutate(pathway = recode(pathway, "GAG" = "Glycosaminoglycan","DF" = "Dietary Fiber", "PG" = "Peptidoglycan", "Resistant_oligosaccharides" = "Resistant oligosaccharides", "Resistant_starch" = "Resistant starch", "Mannans_and_Heteromannans" = "Mannans and \nheteromannans", "Amylose_and_Amylopectin" = "Amylose and \namylopectin", "NSP" = "Non-starch \npolysaccharides"))
```

## Now build up the network graph again, though some steps are not necessary to be repeated as they are already there for the W/NW analysis
```{r}
substrate_significance <- data.frame(
  name = complete_fgsea_results_crc$pathway,
  Value = complete_fgsea_results_crc$padj, ## p-value
  EnrichmentStatus = complete_fgsea_results_crc$NES) ## This is the NES from the FGSEA

# Step 1: Add the values as a vertex attribute in your graph

# Match the node names and assign corresponding values (or NA if no match)
matched <- match(V(g)$name, substrate_significance$name)
V(g)$value_label <- substrate_significance$Value[matched]
V(g)$enrichment_status <- substrate_significance$EnrichmentStatus[matched]
V(g)$is_significant <- !is.na(V(g)$value_label) & V(g)$value_label < 0.05 ## To have a significance label

tg <- as_tbl_graph(g)

layout_df <- tg %>%
  activate(nodes) %>%
  as_tibble() %>%
  group_by(layer) %>%
  mutate(
    x = seq(-1, 1, length.out = n()),
    y = -1 * (layer - 1) * 3  # Increase spacing between layers (e.g., 3 units apart)
  ) %>%
  ungroup()

layout_df <- as_tibble(tg) %>%
  group_by(layer) %>%
  mutate(
    x = seq(-1, 1, length.out = n()) + 1.5,  # shift right
    y = -1.1 * (layer - 1) ## The lower the -x (so closer to) *, the more the white space is reduced
  ) %>%
  ungroup()

coords <- layout_df[, c("x", "y")]
tg <- tg %>% activate(nodes) %>% mutate(x = coords$x, y = coords$y, enrichment_status = enrichment_status)

Fig6C_network_substrate_graph <- ggraph(tg, layout = "manual", x = V(g)$x, y = V(g)$y) +
  geom_mark_rect(aes(x = x, y = y, group = factor(layer)),
                 expand = unit(22, "pt"), color = "white", fill = "lightgrey", alpha = 0.2) +
  #geom_edge_link(edge_width = 0.6, alpha = 0.5, color = "black") + ## Have to play a bit with the edge_with still to get this to work properly
  geom_segment(data = edge_segments, aes(x = x, y = y, xend = xend, yend = yend), color = "gray50", linewidth = 0.6) +
  geom_node_point(data = function(x) dplyr::filter(x, is_significant), shape = 21, size = 8.5, stroke = 0.9, fill = NA, color = "grey20") + ## Draw the ring around the significant nodes
  geom_node_point(aes(color = enrichment_status), size = 8) +
  scale_color_gradient2(low = "#3B6FB6", mid = "lightgrey", high = "#D41646", midpoint = 0, na.value = "lightgrey", name = "NES") +
  # Node names on top
  #geom_node_text(aes(label = name), nudge_y = 0.4, size = 2) + 
  # Node names below
  geom_node_text(aes(label = name), nudge_y = -0.28, size = 1.8) + 
  # Values inside nodes, black color
  geom_node_text(aes(label = ifelse(value_label < 0.05, format(value_label, digits = 2, scientific = TRUE), "")), color = "black", size = 1.5) +
  theme_void() +
  theme(legend.title = element_text(size = 6), legend.text  = element_text(size = 5), legend.key.size = unit(0.65, "lines")) +
  theme(legend.position = c(0.5, 1.15), legend.justification = "center", legend.direction = "horizontal", plot.margin = margin(t = 32,  r = 15, b = 50, l = 15)) + # x = center, y = slightly above top
  coord_cartesian(clip = "off")
```

